First Order Alternation

### 2017-08-16 quantifiers/draft.tex
ok

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
  generalization to more than two sorts is without difficulty, but
  would unnecessarily clutter the technical presentation.} and each
function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
$\#(f)=0$ then $f$ is a \emph{constant}. Let
$\top$ and $\bot$ be constants of sort $\Bool$.

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. A
\emph{term} $t$ of sort $\sigma(t)=S$ is a variable $x \in \vars$
where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{formula} is a quantified
boolean combination of atoms, where the quantifiers are as $\exists x
~.~ \phi$ and $\forall x ~.~ \phi$.

We denote by $\fv{\sigma}{\phi}$ the set of free variables of sort
$\sigma$ in $\phi$ and write $\fv{}{\phi}$ for $\bigcup_{\sigma \in
  \ssorts{\sig}} \fv{\sigma}{\phi}$. For a variable $x \in
\fv{}{\phi}$ and a term $t$ such that $\sigma(t) = \sigma(x)$, let
$\phi[t/x]$ be the result of replacing each occurrence of $x$ by
$t$. For indexed sets $T=\set{t_1,\ldots,t_n}$ and
$X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for
the formula obtained by simultaneously replacing $x_i$ with $t_i$ in
$\phi$, for all $i\in[1,n]$. The size $\len{\phi}$ is the number of
symbols occuring in $\phi$.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
The semantics of quantifiers is given as usual: $\I,\nu \models
\exists x ~.~ \phi$ ($\forall x ~.~ \phi$) if and only if $\I,\nu[x
  \leftarrow u] \models \phi$, for some (for all) value(s) $u \in
\sigma(x)^\I$, where $\nu[x \leftarrow u]$ is the valuation that
assigns $u$ to $x$ and $\nu(y)$ to all $y \neq x$.

@@

Given sets of variables $B,X \subset \vars$ of sort $\Bool$ and
$\Data$, respectively, we denote by $\Form(B,X,Y)$ the set of formulae
$\phi$ such that $\fv{\Bool}{\phi} = B$, $\fv{\Data}{\phi} = X$ and
each variable from $Y$ is of sort $\Data$ and occurs under the scope
of a quantifier. By $\Form^+(B,X,Y)$ we denote the subset of
$\Form(B,X,Y)$ in which each boolean variable occurs under an even
number of negations.

A \emph{first-order alternating data automaton} (fo-ADA or automaton
in the sequel) is a tuple $\A =
\tuple{G,L,Q,\iota,F,\Delta}$, where: \begin{compactitem}
%
\item $G \subseteq \vars$ is a finite set of \emph{global
  variables} of sort $\Data$, 
%
\item $L \subseteq \vars$ is a finite set of \emph{local
  variables} of sort $\Data$,
%
\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
  (\emph{states}),
%
\item $\iota \in \Form^+(Q,\emptyset,\emptyset)$ is the \emph{initial
  configuration},
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{G}\cup G\cup\overline{L},L)$
  is a \emph{transition function},
\end{compactitem}
where, for a set $X \subseteq \vars$, we denote $\overline{X} =
\set{\overline{x} \mid x \in X}$. Intuitively, the variables
$\overline{x}$ from a transition formula $\Delta(q,a)$ refer to past
values, whereas the other variables refer to the current
values. Observe that the initial values are unconstrained (since the
initial configuration is a boolean combination of states without
variables) and quantification is allowed only over the current local
variables. In other words, we do not allow quantification over past
values, which are beyond our control. The size of $\A$ is defined as
$\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

>>>>

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[1,i]$, where $[0]$ is
the empty set.

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
  generalization to more than two sorts is without difficulty, but
  would unnecessarily clutter the technical presentation.} and each
function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
sort $\Data$ and return value $f_\sigma \in \ssorts{\sig}$. If
$\#(f)=0$ then $f$ is a \emph{constant}. Let $\top$ and $\bot$ be
constants of sort $\Bool$ and $\mathbf{0}$ be a constant of sort
$\Data$.

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $x_\sigma$. A
\emph{term} $t$ of sort $t_\sigma = S$ is a variable $x \in \vars$
where $x_\sigma=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $f_\sigma=S$. An
\emph{atom} is a term $p(t_1,\ldots,t_{\#(p)})$ of sort $\Bool$, or an
equality $t \teq s$ between two terms of sort $\Data$. A
\emph{formula} is a possibly quantified boolean combination of atoms.

We denote by $\fv{\sigma}{\phi}$ the set of free variables of sort
$\sigma$ in $\phi$ and write $\fv{}{\phi}$ for $\bigcup_{\sigma \in
  \ssorts{\sig}} \fv{\sigma}{\phi}$. For a variable $x \in
\fv{}{\phi}$ and a term $t$ such that $t_\sigma = x_\sigma$, let
$\phi[t/x]$ be the result of replacing each occurrence of $x$ by
$t$. For indexed sets $T=\set{t_1,\ldots,t_n}$ and
$X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for the formula
obtained by simultaneously replacing $x_i$ with $t_i$ in $\phi$, for
all $i\in[n]$. The size $\len{\phi}$ is the number of symbols
occuring in $\phi$.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow f_\sigma^I$, or an element of
  $f_\sigma^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element of $x_\sigma^\I$. For a term
$t$, we denote by $t^\I_\nu$ the value obtained by replacing each
function symbol $f$ by its interpretation $f^\I$ and each variable $x$
by its valuation $\nu(x)$. For a formula $\phi$, we write $\I,\nu
\models \phi$ if the formula obtained by replacing each term $t$ in
$\phi$ by the value $t^\I_\nu$ is logically equivalent to true.  The
semantics of quantifiers is given as usual: $\I,\nu \models \exists x
~.~ \phi$ if and only if $\I,\nu[x \leftarrow u] \models \phi$, for
some value $u \in x_\sigma^\I$, where $\nu[x \leftarrow u]$ is the
valuation that assigns $u$ to $x$ and $\nu(y)$ to all $y \neq x$.

@@

Given sets of variables $B,X,Y \subset \vars$ of sorts $\Bool$,
$\Data$ and $\Data$, respectively, we denote by $\Form\tuple{B;X;Y}$
the set of formulae $\phi$ such that $\fv{\Bool}{\phi} = B$,
$\fv{\Data}{\phi} = X$ and each variable from $Y$ occurs under the
scope of a quantifier. By $\Form^+\tuple{B;X;Y}$ we denote the subset
of $\Form\tuple{B;X;Y}$ in which each boolean variable from $B$ occurs
under an even number of negations. We denote by $\phi\tuple{B;X;Y}$
any formula in $\Form\tuple{B;X;Y}$.

A \emph{first-order alternating data automaton} (fo-ADA or automaton
in the sequel) is a tuple $\A =
\tuple{G,L,Q,\iota,F,\Delta}$, where: \begin{compactitem}
%
\item $G \subseteq \vars$ is a finite set of \emph{global
  variables} of sort $\Data$, 
%
\item $L \subseteq \vars$ is a finite set of \emph{local
  variables} of sort $\Data$,
%
\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
  (\emph{states}),
%
\item $\iota \in \Form^+\tuple{Q;\emptyset;\emptyset}$ is the \emph{initial
  configuration},
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+\tuple{Q;\overline{G}\cup G\cup\overline{L};L}$ is a
  \emph{transition function},
\end{compactitem}
where, for a set $X \subseteq \vars$, we denote $\overline{X} =
\set{\overline{x} \mid x \in X}$. The \emph{size} of $\A$ is defined
as $\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

Intuitively, the variables $\overline{x}$ from a transition formula
$\Delta(q,a)$ refer to past values, whereas the other variables refer
to the current values. Observe that the initial values of the global
variables are unconstrained (since the initial configuration is a
boolean combination of states without variables) and quantification
occurs only over the current copy of the local variables $L$. As a
remark, the transition relation does not quantify over global
variables $G \cup \overline{G}$, which must be visible throughout the
execution. Furthermore, we assume that the past values of the local
variables are coined and beyond our control, thus quantification over
$\overline{L}$ is not allowed either.

In order to define the semantics of automata, we consider time-stamped
sets of variables $\stamp{X}{n} = \{\stamp{x}{n} \mid x \in X\}$ for
any $n\geq0$. For an input event $a \in \Sigma$, some constant
$n\geq0$ and a formula $\phi$, we write $\stamp{\Delta}{n}(\phi,a)$
for the formula obtained from $\phi$ by simultaneously replacing each
state $q \in \fv{\Bool}{\phi}$ with
$\Delta(\phi,a)[\stamp{G}{n}/\overline{G}, \stamp{G}{n+1}/G,
  \stamp{L}{n}/\overline{L}, \stamp{L}{n+1}/L]$.

Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in
\Sigma[G]^*$ the \emph{run} of $\A$ over $w$ is the sequence of
formulae: \[\iota \stackrel{(a_1,\nu_1)}{\Longrightarrow} \phi_1
\tuple{Q,\stamp{G}{1}, \stamp{L}{1}}
\stackrel{(a_2,\nu_2)}{\Longrightarrow} \ldots
\stackrel{(a_n,\nu_n)}{\Longrightarrow} \phi_n\tuple{Q,\stamp{G}{1}
  \cup \ldots \cup \stamp{G}{n}, \stamp{L}{1} \cup \ldots \cup
  \stamp{L}{n}}\] where $\phi_1 \equiv
\stamp{\Delta}{0}(\iota,a_1)[\mathbf{0}/\stamp{G}{0},\mathbf{0}/\stamp{L}{0}]$
and $\phi_{i+1} \equiv \stamp{\Delta}{i}(\phi_i,a_{i+1})$, for all $i
\in [1,n-1]$. In other words, we assume that all data variables are
initially $\mathbf{0}$.

We say that $\A$ \emph{accepts} $w$ if there exists a valuation $\nu$
such that $\nu(\stamp{x}{i}) = \nu_i(x)$ for each global variable $x
\in G$ and each timestamp $i \in [n]$, $\nu(q) = \true$ for each $q
\in F$, $\nu(q) = \false$ for each $q \in Q \setminus F$ and $\I,\nu
\models \phi_n$. The \emph{language} of $\A$ is the set $L(\A)$ of
words from $\Sigma[G]^*$ accepted by $\A$. In this paper, we ask the
following two questions: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, with $L_1 \cap
  L_2 = \emptyset$ and $Q_1 \cap Q_2 = \emptyset$, do there exist
  automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[G]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}

\subsection{Boolean Closure}

Given a set $B$ of boolean variables, two disjoint sets $X$ and $Y$ of
data variables, for any formula $\phi \in \Form^+\tuple{B;X;Y}$,
define $\overline{\phi} \in \Form^+\tuple{B;X;Y}$ recursively:
\[\begin{array}{rclcrclcrcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} & \hspace*{5mm} & 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} & \hspace*{5mm} & 
\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in B$ is an atom} \\ 
\overline{\exists x ~.~ \phi_1} & \equiv & \forall x ~.~ \overline{\phi_1} && 
\overline{\forall x ~.~ \phi_1} & \equiv & \exists x ~.~ \overline{\phi_1} && 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ is not an atom} \\
&& && && && \overline{\phi} & \equiv & \phi \text{ if $\phi \in B$}
\end{array}\]
Clearly $\phi$ and $\overline{\phi}$ have the same size. The relation
between $\phi$ and $\overline{\phi}$ is formalized below:

\begin{proposition}\label{prop:complement}
  Given sets of variables $B,X,Y \subset \vars$ of sorts boolean, data
  and data respectively, and a formula $\phi \in
  \Form^+\tuple{B;X;Y}$, for any valuation $\nu$ we have $\nu \models
  \phi$ if and only if $\overline{\nu} \models \neg\overline{\phi}$,
  where $\overline{\nu}(x) = \nu(x)$ for all $x \in X$ and
  $\overline{\nu}(b) = \neg\nu(b)$ for all $b \in B$.
\end{proposition}
\proof{By induction on the structure of $\phi$. \qed}

Consider the automata $\A_i = \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$,
for $i=1,2$, where $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 =
\emptyset$. We define the following automata: 
\[\begin{array}{rcl}
\A_\cup & = & \tuple{G,~ L_1 \cup L_2,~ Q_1 \cup Q_2,~ \iota_1 \vee \iota_2,~ F_1 \cup F_2,~ \Delta_1 \cup \Delta_2} \\
\A_\cap & = & \tuple{G,~ L_1 \cup L_2,~ Q_1 \cup Q_2,~ \iota_1 \wedge \iota_2,~ F_1 \cup F_2,~ \Delta_1 \cup \Delta_2} \\
\overline{A_1} & = & \tuple{G,~ L_1,~ Q_1,~ \overline{\iota_1},~ Q_1 \setminus F_1,~ \lambda q \in Q_1~ \lambda a \in \Sigma ~.~ \overline{\Delta_1(q,a)}}
\end{array}\]

\begin{lemma}\label{lemma:boolean-closure}
  Given automata $\A_i = \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$, for
  $i=1,2$, where $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 =
  \emptyset$, we have $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(A_\cap)
  = L(\A_1) \cap L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[G]*
  \setminus L(\A_1)$. 
\end{lemma}
\proof{
  Same as \cite[Lemma 1]{IosifX17}. 
\qed}

\section{Checking Emptiness with Interpolants and Skolem Witnesses}

Since the emptiness problem is undecidable even for automata with only
two global variables and no local variables, by reduction from the
emptiness problem for 2-counter machines \cite{Minsky67}, we give a
semi-algorithm with the following properties. Given an automaton $\A$
as input, the procedure will\begin{inparaenum}[(1)] 
\item return a word $w \in L(\A)$ if $L(\A)
\neq \emptyset$, and 
%
\item if the procedure terminates without returning such a word, then
  we can safely conclude that $L(\A) = \emptyset$. \end{inparaenum}
Due to the undecidability of the emptiness problem, it is possible for
the semi-algorithm to run forever in the case when the language of the
automaton is empty.

\begin{proposition}\label{prop:substitution}
  Given a formula $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$, where $Q_1 \ldots
  Q_n \in \set{\exists,\forall}$ and $\phi$ is a quantifier-free
  formula with free variables $\fv{}{\phi} = \set{x_1,\ldots,x_n,q}$,
  for any formula $\psi$, the following equivalence holds:
  \[Q_1 x_1 \ldots Q_n x_n ~.~ \phi[\psi/q] \iff 
  Q_1 x_1 \ldots Q_n x_n \exists q ~.~ \phi \wedge (q \rightarrow \psi)\]
\end{proposition}
\proof{

\qed}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{abbrv} \bibliography{refs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
